---
title: "List all species present in colorectal cancer datasets"
author: 
  - name: Levi Waldron
    affiliation:
    - CUNY Graduate School of Public Health and Health Policy
    email: levi.waldron@sph.cuny.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('curatedMetagenomicAnalyses')`"
vignette: >
  %\VignetteIndexEntry{Identify all species seen in CRC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This analysis:
1. selects a subset of samples (all CRC-related in this example), 
2. uses the `r CRANpkg("table1")` package to display a table of the characteristics of the included cohort, 
3. sorts species in order of descending prevalence, 
4. uses the `r CRANpkg("DT")` package `datatable` function to display a searchable, paged table of prevalences, and
4. writes the prevalences to file.

Required packages:
```{r}
suppressPackageStartupMessages({
  library(table1)
  library(DT)
  library(curatedMetagenomicData)
  library(dplyr)
})
```

# Select samples

Note that a few species without phylogenetic information are lost.

```{r}
crc_subset <- filter(sampleMetadata, study_condition == "CRC") %>%
  returnSamples(dataType = "relative_abundance",
                rownames = "short")
```

# Cohort characteristics

Create a summary table of the participants in this cohort:

```{r}
table1::table1( ~ disease + disease_subtype + age + gender + country + study_name,
                data = colData(crc_subset))
```

# Show species in order of decreasing prevalence

```{r}
prevalences <- rowSums(assay(crc_subset) > 0) / ncol(crc_subset) 
prevalences <- tibble(species = names(prevalences), prevalence = signif(prevalences, 2)) %>%
  filter(prevalence > 0) %>%
  arrange(-prevalence)
```

```{r}
DT::datatable(prevalences)
```

Write to disk:

```{r}
write.csv(prevalences, row.names = FALSE, file = "prevalences.csv")
```

Download the zipped prevalences file produced by BiocStyle::Biocpkg("curatedMetagenomicData") version 3.2.3: [prevalences.csv.zip](https://github.com/waldronlab/curatedMetagenomicAnalyses/files/8254646/prevalences.csv.zip)

# Session Info

```{r}
sessioninfo::session_info()
```

